Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-05-29 appendix.tex
A bit of confusion :pooh:
add a little more clarification, and fix typos

%Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference; similar arguments hold for the remaining parameters.
We restrict our attention to $x_1,y_1,x_2,y_2\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
Additionally, we exclude the infinite branch length case of $w=0$.
Recall the divergence
\[
\shannonDivergence_{\tau_1,t^*}(\tau_1,\tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \tilde{t}).
\]
By Gibbs inequality, this function attains a unique maximum when
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \tilde{t}) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)
\]
for all $j=1,\ldots,\nSiteSplits$.
Referring to Tables~\ref{tab:gen-sitepatprob} and~\ref{tab:sitepatprob}, we set up simultaneous equations as follows.
The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
\[
-(x^*)^2 = -x_1x_2+y_1w(x_1-x_2)
\]
and for $\{1\}$ and $\{1,2\}$ yields
\[
-(x^*)^2 = -x_1x_2+y_2w(x_2-x_1).
\]
The difference of these two equations simplifies to
\[
0 = (y_1w+y_2w)(x_1-x_2)
\]
and, since $y_1w+y_2w > 0$ by assumption, we have $x_1=x_2$.
This along with the previous equalities imply $x_1=x_2=x^*$ is the unique maximum of the divergence term for $x_1$ and $x_2$.
We use the same process to show that, at the maximum, $y_1=y_2=y^*$, and thus the divergence term attains a unique maximum over $x_1,y_1,x_2,y_2\in(0,1)$ with $w \neq 0$.

@@

\begin{table}
\centering
\begin{tabular}{|l|ll|}
    \hline
    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau_1, \tilde{t})$ & $\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau_1,\tilde{t})$\\
    \hline
     $\{1\}$    &
    $\begin{array}{l}
                    \emptyset\\
                    \{1\}\\
                    \{2\}\\
                    \{1,2\}
                \end{array}$&
    $\begin{array}{l}
                    (1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2)\\
                    (1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2)\\
                    (1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2)\\
                    (1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2)
                \end{array}$\\
    \hline
     $\{1,2\}$    &
    $\begin{array}{l}
                    \emptyset\\
                    \{1\}\\
                    \{2\}\\
                    \{1,2\}
                \end{array}$&
    $\begin{array}{l}
                    (1-x_1)(1+x_2)(1+w)(1-y_1)(1+y_2)\\
                    (1-x_1)(1-x_2)(1-w)(1+y_1)(1-y_2)\\
                    (1+x_1)(1-x_2)(1-w)(1-y_1)(1+y_2)\\
                    (1+x_1)(1-x_2)(1+w)(1+y_1)(1-y_2)
                \end{array}$\\
    \hline
\end{tabular}
\caption{Likelihood calculations for site splits $\{1\}$ and $\{1,2\}$ and ancestral state partitions $\ancestralSplitPartition_j$ on the InvFels tree $\tau_1$ with general branch parameters $\tilde{t}$.
All values multiplied by $1/32$.}
\label{tab:edge-cases}
\end{table}

We now consider the case where there may be more than one ancestral state split that maximizes the partial likelihood for a site split.
Site split $\emptyset$ always attains a maximum at ancestral state split $\emptyset$ provided at least one branch fidelity is not exactly zero.
Table~\ref{tab:edge-cases} gives the possible forms for the partial likelihood for two cases of site splits.
For site split $\{1\}$ we ignore ancestral state split $\{2\}$ as its partial likelihood is always less than the ancestral state split $\emptyset$ except for the case $y_1=y_2=w=0$.
Comparing the remaining ancestral state splits reveals that there is an ambiguity in the form of the likelihood only when either $y_1=y_2=0$, $x_1=x_2=0$, or $w=0$.
We reach the same conclusions doing these comparisons for site splits $\{2\}$, $\{3\}$, and $\{1,2,3\}$.

The remaining site splits $\{1,2\}$, $\{1,3\}$, and $\{2,3\}$ require the following argument.
For site split $\{1,2\}$, we ignore cases where either $y_1=y_2=0$, $x_1=x_2=0$, or $w=0$, and see that an ambiguity in this maximization occurs when
\[
(1-x_1)(1+x_2)(1-y_1)(1+y_2) = (1+x_1)(1-x_2)(1+y_1)(1-y_2),
\]
or, in other words, when the ancestral state splits $\emptyset$ and $\{1,2\}$ both maximize the partial likelihood.
This can happen when $x_1=y_2=u_1$ and $x_2=y_1=u_2$ but $u_1 \neq u_2$, though, when $x_1=x_2=u_1$ and $y_1=y_2=u_2$ the partial likelihood takes the same value.
To resolve this ambiguity, we look to the divergence quanitity that is maximized when $x_1=x_2$ and $y_1=y_2$, and, thus, the full joint likelihood obtains a unique maximum when $x_1=x_2$ and $y_1=y_2$.

>>>>

%Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference; similar arguments hold for the remaining parameters.
We restrict our attention to $x_1,y_1,x_2,y_2\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
Additionally, we exclude the infinite branch length case of $w=0$.
%EM This looks good, but remind us what x^* is here.
%das: fixed
Recall the divergence
\[
\shannonDivergence_{\tau_1,t^*}(\tau_1,\tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \tilde{t}).
\]
where data is generated on the InvFels tree with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$ (Fig.~\ref{fig:farris-fels-top}).
By Gibbs inequality, this function attains a unique maximum when
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \tilde{t}) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)
\]
for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
Referring to Tables~\ref{tab:gen-sitepatprob} and~\ref{tab:sitepatprob}, we set up simultaneous equations as follows.
The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
\[
-(x^*)^2 = -x_1x_2+y_1w(x_1-x_2)
\]
and for $\{1\}$ and $\{1,2\}$ yields
\[
-(x^*)^2 = -x_1x_2+y_2w(x_2-x_1).
\]
The difference of these two equations simplifies to
\[
0 = (y_1w+y_2w)(x_1-x_2)
\]
and, since $y_1w+y_2w > 0$ by assumption, we have $x_1=x_2$.
This along with the previous equalities imply $x_1=x_2=x^*$ is the unique maximum of the divergence term for $x_1$ and $x_2$.
We use the same process to show that the divergence has a unique maximum for $y_1$ and $y_2$ at $y_1=y_2=y^*$, and thus the divergence term attains a unique maximum over $x_1,y_1,x_2,y_2\in(0,1)$ with $w \neq 0$.
This implies the Hessian of $\shannonDivergence$ is negative definite.
%EM IIUC we only use this fact to show that we have a _negative_ second derivative at the ^* parameter values. It would be nice to emphasize that fact, which would also keep people from worrying that we're falling into the "sum of two things with unique optima has a unique optimum."
%das: Ah, right.

@@

\begin{table}
\centering
\begin{tabular}{|l|ll|}
    \hline
    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau_1, \tilde{t})$ & $\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau_1,\tilde{t})$\\
    \hline
     $\{1\}$    &
    $\begin{array}{l}
                    \emptyset\\
                    \{1\}\\
                    \{2\}\\
                    \{1,2\}
                \end{array}$&
    $\begin{array}{l}
    %EM I must be confused here-- these aren't the values I get at all!
    % e.g. for this first one everything is 0 except for the first taxon, which is 1. So shouldn't it be
                   %(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2)\\ ?
    %das: Ha! I must have written out a different case. Went through everything with a fine-tooth comb and noticed one or two slip-ups but everything is basically the same.
                    (1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2)\\
                    (1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2)\\
                    (1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2)\\
                    (1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2)
                \end{array}$\\
    \hline
     $\{1,2\}$    &
    $\begin{array}{l}
                    \emptyset\\
                    \{1\}\\
                    \{2\}\\
                    \{1,2\}
                \end{array}$&
    $\begin{array}{l}
                    (1-x_1)(1+x_2)(1+w)(1-y_1)(1+y_2)\\
                    (1+x_1)(1-x_2)(1-w)(1-y_1)(1+y_2)\\
                    (1-x_1)(1+x_2)(1-w)(1+y_1)(1-y_2)\\
                    (1+x_1)(1-x_2)(1+w)(1+y_1)(1-y_2)
                \end{array}$\\
    \hline
\end{tabular}
\caption{Likelihood calculations for site splits $\{1\}$ and $\{1,2\}$ and ancestral state partitions $\ancestralSplitPartition_j$ on the InvFels tree $\tau_1$ with general branch parameters $\tilde{t}$.
All values multiplied by $1/32$.}
\label{tab:edge-cases}
\end{table}

We now consider the case where there may be more than one ancestral state split that maximizes the partial likelihood for a site split.
Site split $\emptyset$ always attains a maximum at ancestral state split $\emptyset$ provided at least one branch fidelity is not exactly zero.
Table~\ref{tab:edge-cases} gives the possible forms for the partial likelihood for two cases of site splits.
For site split $\{1\}$ we ignore ancestral state split $\{2\}$ as its partial likelihood is always less than the ancestral state split $\emptyset$ except for the case $y_1=y_2=w=0$.
Comparing the remaining ancestral state splits reveals that there is an ambiguity in the form of the likelihood only when one of $x_1,y_1,x_2,y_2,$ or $w$ is equal to zero, which are cases we exclude.
For example, in comparing ancestral state splits $\emptyset$ and $\{1\}$ an ambiguity arises when
\[
(1-x_1)(1+x_2)(1+w) = (1+x_1)(1-x_2)(1-w),
\]
which holds either if $x_2=0$ and $w=x_1$ or $w=0$ and $x_1=x_2$.
We reach the same conclusions doing these comparisons for site splits $\{2\}$, $\{3\}$, and $\{1,2,3\}$.

The remaining site splits $\{1,2\}$, $\{1,3\}$, and $\{2,3\}$ require the following argument.
For site split $\{1,2\}$, we ignore cases where any of $x_1,y_1,x_2,y_2,$ or $w$ are zero and see that an ambiguity in this maximization can occur when
\[
(1-x_1)(1+x_2)(1-y_1)(1+y_2) = (1+x_1)(1-x_2)(1+y_1)(1-y_2),
\]
or, in other words, when the ancestral state splits $\emptyset$ and $\{1,2\}$ both maximize the partial likelihood.
This can happen when $x_1=y_2=u_1$ and $x_2=y_1=u_2$ but $u_1 \neq u_2$, though, when $x_1=x_2=u_1$ and $y_1=y_2=u_2$ the partial likelihood takes the same value.
%das: @EM does the below make sense? I know you were worried earlier about arguing the max of the sum of two things, but I think to resolve this ambiguity it's necessary. Basically, we want to show we get a unique t from (max_t f(t) + (max_h g(t,h))) for t branch lengths and h ancestral states. To do this, if max_h g(t,h) has more than one value we can look at f(t) + max_h g(t,h). I.e., max_h g(t,h) happens if either {x_1=y_2,x_2=y_1} or {x_1=x_2,y_1=y_2}, but f(t)|_{x_1=x_2,y_1=y_2} > f(t)|_{x_1=y_2,x_2=y_1} (where |_{} is "restricted to") so the full likelihood will have a unique maximum, and that's what we're concerned about.
To resolve this ambiguity, we look to the divergence quanitity that is maximized when $x_1=x_2$ and $y_1=y_2$, and, thus, the full joint likelihood obtains a unique maximum when $x_1=x_2$ and $y_1=y_2$.
It is possible that ancestral state splits $\{1\}$ and $\{2\}$ both maximize the partial likelihood as well, but this case follows a similar argument.
